---
title: "Text Manipulation"
author: "Statistical Computing, 36-350"
date: "Tuesday October 4, 2022"
---

Last week: Dplyr, pipes, and more
===

- The tidyverse is a collection of packages for common data science tasks
- Tidyverse functionality is greatly enhanced using pipes (`%>%` operator)
- Pipes allow you to string together commands to get a flow of results
- `dplyr` is a package for data wrangling, with several key verbs (functions)
- `filter()`: subset rows based on a condition
- `group_by()`: define groups of rows according to a condition
- `summarize()`: apply computations across groups of rows
- `arrange()`: order rows by value of a column
- `select()`: pick out given columns
- `mutate()`: create new columns
- `mutate_at()`: apply a function to given columns
- `tidyr` is a package for manipulating the structure of data frames
- `pivot_longer()`: make "wide" data longer
- `pivot_wider()`: make "long" data wider

<span style="color: blue">Part I</span>
===

<span style="color: blue">*String basics*</span>

What are strings?
===

The simplest distinction:

- **Character:** a symbol in a written language, like letters, numerals, punctuation, space, etc.

- **String:** a sequence of characters bound together

```{r}
class("r")
class("Ryan")
```

Why do we care about strings?

- A lot of interesting data out there is in text format!
- Webpages, emails, surveys, logs, search queries, etc.
- Even if you just care about numbers eventually, you'll need to understand how to get numbers from text

Whitespaces
===

Whitespaces count as characters and can be included in strings:

- `" "` for space
- `"\n"` for newline
- `"\t"` for tab

```{r}
str = "Dear Mr. Carnegie,\n\nThanks for the great school!\n\nSincerely, Ryan"
str
```

Use `cat()` to print strings to the console, displaying whitespaces properly

```{r}
cat(str)
```

Vectors/matrices of strings
===

The character is a basic data type in R (like numeric, or logical), so we can make vectors or matrices of out them. Just like we would with numbers

```{r}
str.vec = c("Statistical", "Computing", "isn't that bad") # Collect 3 strings
str.vec # All elements of the vector
str.vec[3] # The 3rd element
str.vec[-(1:2)] # All but the 1st and 2nd

str.mat = matrix("", 2, 3) # Build an empty 2 x 3 matrix
str.mat[1,] = str.vec # Fill the 1st row with str.vec
str.mat[2,1:2] = str.vec[1:2] # Fill the 2nd row, only entries 1 and 2, with
                              # those of str.vec
str.mat[2,3] = "isn't a fad" # Fill the 2nd row, 3rd entry, with a new string
str.mat # All elements of the matrix
t(str.mat) # Transpose of the matrix
```

Converting other data types to strings
===

Easy! Make things into strings with `as.character()`

```{r}
as.character(0.8)
as.character(0.8e+10)
as.character(1:5)
as.character(TRUE)
```

Converting strings to other data types
===

Not as easy! Depends on the given string, of course

```{r}
as.numeric("0.5")
as.numeric("0.5 ")
as.numeric("0.5e-10")
as.numeric("Hi!")
as.logical("True")
as.logical("TRU")
```

Number of characters
===

Use `nchar()` to count the number of characters in a string

```{r}
nchar("coffee")
nchar("code monkey")
length("code monkey")
length(c("coffee", "code monkey"))
nchar(c("coffee", "code monkey")) # Vectorization!
```

<span style="color: blue">Part II</span>
===

<span style="color: blue">*Substrings, splitting and combining strings*</span>


Getting a substring
===

Use `substr()` to grab a subsequence of characters from a string, called a **substring**

```{r}
phrase = "Give me a break"
substr(phrase, 1, 4)
substr(phrase, nchar(phrase)-4, nchar(phrase))
substr(phrase, nchar(phrase)+1, nchar(phrase)+10)
```

`substr()` vectorizes
===

Just like `nchar()`, and many other string functions

```{r}
presidents = c("Clinton", "Bush", "Reagan", "Carter", "Ford")
substr(presidents, 1, 2) # Grab the first 2 letters from each
substr(presidents, 1:5, 1:5) # Grab the first, 2nd, 3rd, etc.
substr(presidents, 1, 1:5) # Grab the first, first 2, first 3, etc.
substr(presidents, nchar(presidents)-1, nchar(presidents)) # Grab the last 2
                                                           # letters from each
```

Replace a substring
===

Can also use `substr()` to replace a character, or a substring

```{r}
phrase
substr(phrase, 1, 1) = "L"
phrase # "G" changed to "L"
substr(phrase, 1000, 1001) = "R"
phrase # Nothing happened
substr(phrase, 1, 4) = "Show"
phrase # "Live" changed to "Show"
```

Splitting a string
===

Use the `strsplit()` function to split based on a keyword

```{r}
ingredients = "chickpeas, tahini, olive oil, garlic, salt"
split.obj = strsplit(ingredients, split=",")
split.obj
class(split.obj)
length(split.obj)
```

Note that the output is actually a list! (With just one element, which is a vector of strings)

`strsplit()` vectorizes
===

Just like `nchar()`, `substr()`, and the many others

```{r}
great.profs = "Nugent, Genovese, Greenhouse, Seltman, Shalizi, Ventura"
favorite.cats = "tiger, leopard, jaguar, lion"
split.list = strsplit(c(ingredients, great.profs, favorite.cats), split=",")
split.list
```

- Returned object is a list with 3 elements
- Each one a vector of strings, having lengths 5, 6, and 4 
- Do you see why `strsplit()` needs to return a list now?

Splitting character-by-character
===

Finest splitting you can do is character-by-character: use `strsplit()` with `split=""`

```{r}
split.chars = strsplit(ingredients, split="")[[1]]
split.chars
length(split.chars)
nchar(ingredients) # Matches the previous count
```

Combining strings
===

Use the `paste()` function to join two (or more) strings into one, separated by a keyword

```{r}
paste("Spider", "Man") # Default is to separate by " "
paste("Spider", "Man", sep="-")
paste("Spider", "Man", "does whatever", sep=", ")
```

`paste()` vectorizes
===

Just like `nchar()`, `substr()`, `strsplit()`, etc. Seeing a theme yet?

```{r}
presidents
paste(presidents, c("D", "R", "R", "D", "R"))
paste(presidents, c("D", "R")) # Notice the recycling (not historically accurate!)
paste(presidents, " (", 42:38, ")", sep="")
```

Condensing a vector of strings
===

Can condense a vector of strings into one big string by using `paste()` with the `collapse` argument

```{r}
presidents
paste(presidents, collapse="; ")
paste(presidents, " (", 42:38, ")", sep="", collapse="; ")
paste(presidents, " (", c("D", "R", "R", "D", "R"), 42:38, ")", sep="", collapse="; ")
paste(presidents, collapse=NULL) # No condensing, the default
```

<span style="color: blue">Part III</span>
===

<span style="color: blue">*Reading in text, summarizing text*</span>

Text from the outside
===

How to get text, from an external source, into R? Use the `readLines()` function

```{r}
king.lines = readLines("https://www.stat.cmu.edu/~arinaldo/Teaching/36350/F22/data/king.txt")
class(king.lines) # We have a character vector
length(king.lines) # Many lines (elements)!
king.lines[1:3] # First 3 lines
```

(This was Martin Luther King Jr.'s famous "I Have a Dream" speech at the March on Washington for Jobs and Freedom on August 28, 1963)

Reading from a local file
===

We don't need to use the web; `readLines()` can be used on a local file. The following code would read in a text file from Professor Rinaldo's computer:

```{r, error=TRUE}
king.lines.2 = readLines("~/Dropbox/Teaching/36-350/36-350_F22/webpage/data/king.txt")
```

This will cause an error for you, unless your folder is set up exactly like Professor Rinaldo's laptop! So using web links is more robust

Reconstitution
===

Fancy word, but all it means: make one long string, then split the words

```{r}
king.text = paste(king.lines, collapse=" ")
king.words = strsplit(king.text, split=" ")[[1]]

# Sanity check
substr(king.text, 1, 150)
king.words[1:20]
```

Counting words
===

Our most basic tool for summarizing text: **word counts**, retrieved using `table()`

```{r}
king.wordtab = table(king.words)
class(king.wordtab)
length(king.wordtab)
king.wordtab[1:10]
```

What did we get? Alphabetically sorted unique words, and their counts = number of appearances

The names are words, the entries are counts
===

Note: this is actually a vector of numbers, and the words are the names of the vector

```{r}
king.wordtab[1:5]
king.wordtab[2] == 2
names(king.wordtab)[2] == "-"
```

So with named indexing, we can now use this to look up whatever words we want

```{r}
king.wordtab["dream"] 
king.wordtab["Negro"]
king.wordtab["freedom"]
king.wordtab["equality"] # NA means King never mentioned equality
```

Most frequent words
===

Let's sort in decreasing order, to get the most frequent words

```{r}
king.wordtab.sorted = sort(king.wordtab, decreasing=TRUE)
length(king.wordtab.sorted)
head(king.wordtab.sorted, 20) # First 20
tail(king.wordtab.sorted, 20) # Last 20
```

Notice that punctuation matters, e.g., "York" and "York." are treated as separate words, not ideal---we'll learn just a little bit about how to fix this on lab, using **regular expressions**

Visualizing frequencies
===

Let's use a plot to visualize frequencies

```{r}
nw = length(king.wordtab.sorted)
plot(1:nw, as.numeric(king.wordtab.sorted), type="l",
     xlab="Rank", ylab="Frequency")
```

A pretty drastic looking trend! It looks as if $\mathrm{Frequency} \propto (1/\mathrm{Rank})^a$ for some $a>0$

Zipf's law
===

This phenomenon, that frequency tends to be inversely proportional to a power of rank, is called **Zipf's law** 

For our data, Zipf's law approximately holds, with $\mathrm{Frequency} \approx C(1/\mathrm{Rank})^a$ for $C=100$ and $a=0.65$

```{r}
C = 100; a = 0.65
king.wordtab.zipf = C*(1/1:nw)^a
cbind(king.wordtab.sorted[1:8], king.wordtab.zipf[1:8])
```

Not perfect, but not bad. We can also plot the original sorted word counts, and those estimated by our formula law on top

```{r}
plot(1:nw, as.numeric(king.wordtab.sorted), type="l",
     xlab="Rank", ylab="Frequency")
curve(C*(1/x)^a, from=1, to=nw, col="red", add=TRUE)
```

We'll learn about plotting tools in detail a bit later

Summary
===

- Strings are, simply put, sequences of characters bound together
- Text data occurs frequently "in the wild", so you should learn how to deal with it!
- `nchar()`, `substr()`: functions for substring extractions and replacements
- `strsplit()`, `paste()`: functions for splitting and combining strings
- Reconstitution: take lines of text, combine into one long string, then split to get the words
- `table()`: function to get word counts, useful way of summarizing text data
- Zipf's law: word frequency tends to be inversely proportional to (a power of) rank